### 2017-02-21 2017mergingmerges/section/XX-Fusion.tex
process evaluation

The core fusion algorithm works by constructing a static schedule for a pair of processes, which is repeated over the whole network until only one process remains.
The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.

@@

Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
Function \ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
The order processes are fused can affect whether fusion succeeds or fails, so all permutations are tried.
We address this in more detail in~\S\ref{s:FusionOrder}.

@@

Function \ti{tryStepPair} defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.

@@

Figure~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.

@@

Figure~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
\ti{channels} computes the $\ChanType_2$ map for a pair of processes.
\ti{connected} checks whether two processes have any shared inputs or outputs - basically whether it is worth fusing them.
\ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
\ti{swaplabels} flips the order of the compound labels in an instruction.

>>>>

The core fusion algorithm works by constructing a static schedule for a pair of processes, which is repeated over the whole network until only one process remains.
The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.

@@

Fig.~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
Function \ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
The order processes are fused can affect whether fusion succeeds or fails, so all permutations are tried.
We address this in more detail in~\S\ref{s:FusionOrder}.

@@

Function \ti{tryStepPair} defined in Fig.~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.

@@

Fig.~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.

@@

Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
\ti{channels} computes the $\ChanType_2$ map for a pair of processes.
\ti{connected} checks whether two processes have any shared inputs or outputs - basically whether it is worth fusing them.
\ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
\ti{swaplabels} flips the order of the compound labels in an instruction.

### 2017-02-21 2017mergingmerges/section/XX-Processes.tex
process evaluation

The grammar for processes is shown in figure~\ref{fig:Process:Def}.
Channels, labels and variables are specified by some external, globally unique set of names.
For values and expressions we use a simple untyped lambda calculus with a few primitives chosen to facilitate the examples.

\amos{Settle on a terminology for channels / streams. Are both necessary?}
Streams are the abstract data flowing through while channels are particular endpoints, so a stream can have multiple channels.
A stream can have at most one output channel, and any number of input channels.

A process defines a stream computation, taking any number of input streams and producing at least one output stream.
The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
The input states are explained in detail later in \S\ref{s:Process:Eval}.

@@

The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

As usual with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer process to push to that stream.

All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
% Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.

A process network is a set of multiple processes that can be evaluated concurrently.
Any inputs that are not produced as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

We now describe evaluation of processes and process networks.
We split evaluation into three main parts:
\begin{itemize}
\item Injection, where values are inserted into a process' input buffer.
Injection is only possible when the process input buffer is empty.
\item Shaking, where a process takes a step from one label to another.
Shaking a process results in an updated process as well as an output message.
If a process pushes to a stream, the push value must be able to be injected to other processes.
\item Feeding, where an environment of input values are fed to the processes, and output values are collected.
This is the `top level' of evaluation that uses both injection and shaking.
\end{itemize}

@@

Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.

Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall network's input streams.
Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.

Injection can only happen when a process is ready to receive more input.
A process has a single element buffer for each input, stored in its input state.
This can be either @none@ meaning an empty buffer, @pending@ meaning a single value has been added to the buffer but has not been read yet, or @have@ meaning the value was added and in the process of being used.

(InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded (or at least multiple element) buffer.
Injecting the value puts the value as @pending@ in the buffer.

(InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.

(ProcessesInject) performs injection over a process network.
Every process in the network must have the value injected into it.
This means if multiple processes read from that stream, all input buffers for that stream must be empty.

The rules in figure~\ref{fig:Process:Eval:Shake} are the `shake' part of evaluation, where actual computation occurs. 
As usual, $\alpha$ denotes the message type, with $\tau$ being an internal message. The \Push~ message is a single value being output on a channel.

The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
The output message $\alpha$ can be an internal state change or an emitted value.
The result also has the new label, the new input state buffers, and the substitution to apply to the heap.

The two judgment forms for shaking processes are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p}}$.
The process shaking just shakes a single instruction and updates the process.
Shaking a process network chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the network.

(Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.

(Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull.

(Push) evaluates the push expression $e$ under the heap, and sends the value as the message.

(Jump) simply returns the new label and substitution.

(CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.

(Shake) unwraps a single process and evaluates the instruction.
The instruction updates are evaluated and updated in the process heap.
It updates the process with the new label, input state and heap.

(ProcessesInternal) chooses one process from the network and evaluates it.
When the process evaluates with an internal message ($\tau$), the entire network evaluates by replacing that process.

(ProcessesPush) chooses one process from the network and evaluates it, where the process evaluates with a push message.
The emitted push message is then injected into all other processes in the network, which means they must either ignore the channel or be ready to add it to their buffer.
If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
The entire network then emits the same message.

The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process network and output values are accumulated.
The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
Feeding evaluates the process network until all input values have been injected.

@@

(FeedStart) is the axiom form where all input values have been injected and there are no input values left.
This is the start of evaluation.

(FeedInternal) first recursively feeds its input accumulator and process network, then allows the resulting network to take an internal step.
The internal step does not affect the accumulators.
% The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.

(FeedPush) works similarly to (FeedInternal) except that the process network emits a push message.
The pushed value is added to the end of the accumulator for that channel.

(FeedExternal) allows inputs to be injected into the process network.
For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
The recursive feed is evaluated with the last value removed from the accumulators.
The last value is then injected into the network, and added back to the result accumulators.

>>>>

The grammar for processes is shown in Fig.~\ref{fig:Process:Def}.
Channels, labels and variables are specified by some external, globally unique set of names.
For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples.

We refer to the \emph{endpoint} of a stream as a channel.
A stream may flow through many input channels on different processes, but can only be produced from at most one output channel.

Each input channel is paired with an input state, which can be either @none@ (no value), @have@ (a value has been pulled), or ($@pending@~\Value$) (a value is available to be pulled).
These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
These should be initialised to @none@ for a new process.
Input states are explained in detail later in \S\ref{s:Process:Eval}.

@@

The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

As usual with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.

All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.

A process network is a set of multiple processes that can be evaluated concurrently.
Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

Evaluation of processes and process networks is split into three parts:
\begin{itemize}
\item Injection, where a single value for a stream is inserted into a process' input buffer.
Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
Injection for a process network proceeds by injecting the same value into all processes in the network.

\item Shaking, where a process takes a step from one label to another.
Shaking a process results in an updated process and an optional pushed value.
When a process pushes to a channel, the value is injected to other processes.
Shaking a process network chooses any process that can be shaked, and updates it in the process network.

\item Feeding, where an environment of input values are fed to the process network, and output values are collected.
Feeding alternates between injecting and shaking: injecting values from the environment when processes are ready to receive values, and repeatedly shaking the processes to collect their outputs.
\end{itemize}

@@

% Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.

\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} defines the judgment forms and rules for injection.
Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.

A process has a single element buffer for each input, stored in its input state.
This can be either @none@, in which case the buffer is empty and a value may be injected; @pending@, where a single value has been injected to the buffer but has not been pulled yet; or @have@, where a value has already been injected and is currently being used.

The judgment form ($\ProcInject{p}{c}{v}{p'}$) means that the process $p$ has the value $v$ inserted to the input buffer for channel $c$, resulting in the updated process $p'$.

Rule (InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded, or at least multiple element, buffer.
We use the syntax ($\HeapUpdateOne{@ins@~p}{c}{@pending@~v}$) to perform a map update, in this case removing the ($c=@none@$) from ($@ins@~p$) and then inserting ($c=@pending@~v$).
This sets the input buffer as pending, meaning that any future attempts to inject to this channel will fail, until the process pulls and then drops the value.

Rule (InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.

Rule (ProcessesInject) performs injection over a process network.
Every process in the network must have the value injected into it.
This means if multiple processes read from that stream, all input buffers for that stream must be empty.

\subsubsection{Shaking}
Fig.~\ref{fig:Process:Eval:Shake} defines the judgment forms and rules for shaking a process, allowing it to take a step, and produce an output message.
% I don't like them, but alpha and tau are standard names in process calculi. 
The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.

The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
The result contains an output message $\alpha$, as well as the new label $l'$, the new input state $i'$, and the update substitution to apply to the heap $u'$.

Rule (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.

Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.

Rule (Push) evaluates the push expression $e$ under the heap, and sends that value as output the message.

Rule (Jump) simply returns the new label and substitution.

Rules (CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.

The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
The process shaking shakes a single instruction and updates the process.
Shaking a process network chooses a single process to shake and updates it in the network.
If shaking the process resulted in an output value, that value is injected into all the other processes in the network.

Rule (Shake) takes a single process and shakes its instruction.
The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
The process steps to the new label with the new input state, and the values are updated in the heap.

Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces an internal message ($\tau$).

Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces a push message.
The emitted push message is then injected into all other processes in the network, meaning they must either ignore the channel or be ready to receive it in their buffer.
If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
The entire network then emits the same message.

\subsubsection{Feeding}
Fig.~\ref{fig:Process:Eval:Feed} defines the judgment forms and rules for feeding, where external input values are fed into a process network and output values are accumulated.
The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
Feeding evaluates the process network until all input values have been injected.

@@

Rule (FeedStart) applies when all input values have been injected and there are no input values left.
In this case, the output values are the same as the input values.

Rule (FeedInternal) allows the process network to take an internal step.
It first feeds its input accumulator and process network, then allows the resulting network to take an internal step.

Rule (FeedPush) allows the process network to emit a push message.
As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message.
The pushed value is collected in the accumulator list for that stream.

Rule (FeedExternal) allows inputs to be injected into the process network.
For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
The recursive feed is evaluated with the last value removed from the accumulators.
The last value is then injected into the network, and added back to the result accumulators.
